Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-07-27 Manuscript/proto-manuscript.tex
There was a manuscript section accidentaly erased This cherry-pick merge restores it (Sim details) from commit MS: wrote section Simulation details

\begin{abstract}
  % insert abstract here
  \todo{Abstract directly from Joe's conference abstracts. To be rewritten.}
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. 
However, there is still a room for improvements in current force fields – it is known from the literature,
that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. 
We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the
lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent
ions yielding models that feature correct ion pairing [3]. 
Using the electrometer concept [3, 4] and x-ray scattering form factors, our simulations point out that our hypothesis is
correct and ECC is indeed a missing important contribution in current classical lipid models. 
Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECC-lipids", shows accurate binding affinity to sodium and calcium cations and headgroup order parameter response to bound charge. 
We also provide for the first time a realistic stochiometry of bound calcium cations to a POPC membrane, and their binding sites. 
This work will continue as an open collaboration project NMRlipids IV (http://nmrlipids.blogspot.fi).
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
Cation interactions with lipid bilayers serving as a model for cellular
membranes are thus widely studied with experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{Berkowitz12,??}. General conclusion from experimental
studies has been that multivalent ions and lithium have weak specific binding in
phopholipid bilayers, while other monovalent ions do not essentially
bind \cite{cevc90,tocanne90,seelig90,binder02}. The presence of anoinic lipids, like PS or PG,
increase the concentration close to the bilayer and thus the amount of bound
ions, but do not affect the specific binding constant \cite{seelig90}.
The binding details, like binding sites and stoichiometry are not yet fully
resolved but interpretation of NMR and scattering experiments suggest that one
Ca2+ interacts mainly with the choline groups \cite{hauser76,hauser78,herbette84} of two
phospholipid molecules \cite{altenbach84}.

Classical molecular dynamics simulations have potential to solve the ion binding details
in lipid bilayers and reveal its relevance to various biological
problems \cite{bockmann03,bockmann04,??}. However, the available
molecular dynamics simulation models have a strong tendency to overestimate
cation binding on zwitterionic bilayers and none of them can reproduce the
experimental data with the accuracy required for the interpretation of
experiments \cite{catte16}. The overestimated specific cation binding also
makes zwitterionic bilayers effectively positively charged, which could
potentially lead to significant artefacts in applications.
Thus, there is a vast demand for classical molecular dynamics simulation
model which correctly reproduces cation binding in lipid bilayers.

The lack of electronic polarizability from the classical MD simulation
models is a potential source of artefact, which could lead to overbinding
of cations. 
%Almost all lipid bilayer simulations reported in the literature 
%are done with force fields that represent interactions between atoms
%with pairwise additive empirical potential and exclude the electronic 
%polarizability \cite{SOME_review, FF_papers}.
The issue has been considered higly relevant since the early days of
lipid bilayer simulations and pioneering simulation studies scaled
the partial charges to effectively include electronic
polarizability \cite{jonsson86,egberts94}. The explicit inclusion of
electronic polarizability has been, however, turned to be 
practically complicated and implementations for lipids are rare \cite{chowdhary13}.

In this work we show that the cation overbinding in classical MD simulations
can be corrected by including electronic polarizability by using effective continuum
correction (ECC)~\cite{leontyev11} for polar region of zwitterionic lipid molecules.
This is essentially physicalle well justified version of partial charge scaling
implemented in early days of lipid and surfctant simulations \cite{jonsson86,egberts94}. 
The approach has been previously shown in to improve bulk performace of
ion models against neutron scattering data \cite{Jungwirth2015,kohagen14,kohagen16}.
The better bulk behaviour was not, however, sufficient to correct binding in lipid
bilayers \cite{catte16}. 
%motivation for its application to zwitterionic lipids like POPC.

@@

\subsection{Electronic continuum correction for lipid bilayers}
According to the electronic continuum correction (ECC)\cite{leontyev11}, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$. 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}. % that exists in living cells. 
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

Here we apply the electronic continuum correction to lipid bilayers to accurately 
describe the lipid headgroup response to Na+ and Ca2+ concentrations \cite{catte16}. 
We use the Lipid14~\cite{dickson14} force field parameters as a starting point,
because they give the most realistic headgroup response with added cations 
and relatively realistic glycerol backbone and headgroup structures
when compared with other state of the art lipid models \cite{botan15,catte16}.
The partial atomic charges in Lipid14 were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum.
If the charges are obtained by using RESP in an implicit solvent, 
they vary the most for the polar moieties \cite{maciejewski14}. 
In order to implicitly capture solvent induced polarization 
in models with fixed charges, 
we shall use the average of partial charges 
from both environments (i.e. vaccum and solvent),
so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway, I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

Hydrocarbon chains in Lipid14 and other lipids models are highly optimized and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16},
in contrast to glycerol backbone and headgroup regions which require
some improvement in all available lipid models \cite{botan15}. 
To minimize the detuning of the highly optimized hydrocarbon
parameters, we apply ECC only
to the headgroup, glycerol backbone and carbonyl in acyl chains.
These regions are also the most polar parts in lipids and are expected
to most affect cation binding.   
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications.

Mere scaling of partial charges in the modified region by the factor $f_q$ reduces the 
area per molecule to ??\todo{find the value}, which is significantly smaller than the
experimental value (\cite{}) and the original Lipid14 values (\cite{})\todo{Add the values}.  
The decrease of area was speculated 
\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
to arise from reduced hydration
of headgroup due to the lower polarity of molecules with scaled charges.
The hydration can be increased by decreasing the radius of atoms
by reducing the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
This allows water molecules to approach closer to lipid atoms and
have stonger electrostatic interactions with them.
\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Indeed, by reducing $\sigma$ with a scaling factor $f_\sigma = 0.89$ 
we increased the area per molecule
to a level close to experimental and original Lipid14 values. 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison to experimental data}
Structures sampled by individual lipid molecules in simulations were compared
to experimental data by using C-H bond order parameters \cite{ollila16}
\begin{equation}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between C-H bond and membrane
normal and average is taken over all sampled configurations.

Bilayer dimensions were compared to experiments by using the
scattering form factor \cite{ollila16}
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}. More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}.

@@

@@

\todo{To be written}

\subsubsection{Simulations with cationic surfactants}
Automated topoly builder \cite{malde11} was first used to create pdb structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
The AmberTools \cite{amber} was then used to generate the Amber force field
parameters from the pdb file. These were converted to Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond the corresponding segments in Lipid14 \cite{dickson14}.
The surfactants were randomly mized with lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfctant in POPC bilayer.
Simulation parameters and force field for POPC was already used and
described previously \cite{catte16} and were downloaded from \cite{lipid14POPC0mMNaClfiles}.
The systems were simulated 100-200~ns at 313K and the first 20~ns of the trajectories
were excluded from the analysis as an equilibration period.
The simulations were ran with Gromacs 5~\cite{abraham15}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

\subsection{Lipid bilayers without ions}
The scattering form factors, NMR order parameters and area per lipids calculated from 
the ECC corrected lipid model for POPC are compared to experiments and original
Lipid14 results in Fig. \ref{simVSexpNOions} and in Table~\ref{tab:apls}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
The structural quality is comparable to the state of art lipid models available in literature \cite{ollila16},
thus we conclude that the ECC corrected lipid model reproduces the lipid bilayer structure
in liquid disordered phase with similar accuracy than other available models
\todo{Discussion to be finished when we have all the results in the figure}.
\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
%  \caption{\label{fig:form-f}
%    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models.  }
%\end{figure}
%
%\begin{figure}[]
%  \centering
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 and ECCLipids17 models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and EECLipid17 models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{Add acyl chain order parameters, POPC chemical structure}
  \todo{Should we add the results from original lipid14?}
\end{figure}
%Area per molecules extracted from MD simulations and SPD model fitted to scattering data \todo{check that this the case for the used values}
%are shown in Table \ref{tab:apls}.
\todo{finalize figure (NMR headgr. OPs + SAXS, continue in the discussion.}
\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{tab:apls} }
  \begin{tabular}{c c c}
    model          & A (Å$^2$)   & Temperature [K] \\
    \hline
    Lipid14 (literature)  & 65.6$\pm$ 0.5  &  303 \\
    Lipid14ecc0.80+sigma0.875 &        &  313    \\
    GMX small patch           & 64.9   &         \\
    GMX 4xbig patch           & 65.5   &         \\
    oMM small patch           & 63.65  &         \\
    oMM 4xbig patch           & 63.7   &         \\
    \hline
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.}  & 62.7  &  293    \\
    experiment  & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    experiment  & 68.1  &  333    \\
    experiment POPE  & 56.6 &  303    \\
    \hline
  \end{tabular}
\end{table}

>>>>

\begin{abstract}
  % insert abstract here
  \todo{Abstract directly from Joe's conference abstracts. To be rewritten.}
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. 
However, there is still a room for improvements in current force fields – it is known from the literature,
that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. 
We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the
lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent
ions yielding models that feature correct ion pairing [3]. 
Using the electrometer concept [3, 4] and x-ray scattering form factors, our simulations point out that our hypothesis is
correct and ECC is indeed a missing important contribution in current classical lipid models. 
Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECC-lipids", shows accurate binding affinity to sodium and calcium cations and headgroup order parameter response to bound charge. 
We also provide for the first time a realistic stochiometry of bound calcium cations to a POPC membrane, and their binding sites. 
This work will continue as an open collaboration project NMRlipids VI (http://nmrlipids.blogspot.fi).
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
Zwitterionic phosphocholine (PC) lipid bilayers have been widely used as model systems
to understand molecular level details of specific cation interactions with
cellular membranes by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??} \todo{We need to select references for these}.
While relative binding affinity of different ions is agreed to follow Hoffmeister
serie, the molecular details of binding and binding energetics are
not fully understood \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding model and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.

%The binding details, like binding sites and stoichiometry are not yet fully
%resolved but interpretation of NMR and scattering experiments suggest that one
%Ca2+ interacts mainly with the choline groups \cite{hauser76,hauser78,herbette84} of two
%phospholipid molecules \cite{altenbach84}.

In this work we show that the accuracy of Calcium binding
on zwitterionic PC lipid bilayer can be significantly improved by including electronic polarizability
by using effective continuum correction (ECC)~\cite{leontyev11} for polar region of zwitterionic lipid molecules
in classical MD simulation model.
This is essentially physically well justified version of partial charge scaling
implemented in early days of lipid and surfactant simulations \cite{jonsson86,egberts94}. 
The approach has been previously shown in to improve bulk performace of
ion models against neutron scattering data \cite{Jungwirth2015,kohagen14,kohagen16}.
The better bulk behaviour was not, however, sufficient to correct binding in lipid
bilayers \cite{catte16}. 
%motivation for its application to zwitterionic lipids like POPC.

%Classical molecular dynamics simulations have potential to solve the ion binding details
%in lipid bilayers and reveal its relevance to various biological
%problems \cite{bockmann03,bockmann04,??}. However, the available
%molecular dynamics simulation models have a strong tendency to overestimate
%cation binding on zwitterionic bilayers and none of them can reproduce the
%experimental data with the accuracy required for the interpretation of
%experiments \cite{catte16}.

The proposed MD simulation model with ECC-corrected lipids and ions
reproduce the experimental NMR data of Calcium binding in PC lipid bilayers
and mainly supports the proposed ternary complex model (2 Ca2+ bind in
on POPC) \cite{akutsu81,altenbach84}. However, non-negligible contribution
of one Ca2+ cross bridging three POPC molecules is also observed.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions. Thus, the proposed approach to improve the accuracy of cation
binding in current MD models is highly useful for future MD simulations
with physiological salt conditions.

%re is a vast demand for classical molecular dynamics simulation
%model which correctly reproduces cation binding in lipid bilayers.

@@

\subsection{Electronic continuum correction for lipid bilayers}
The lack of electronic polarizability from the classical MD simulation
models has been considered higly relevant issue since the early days of
lipid bilayer simulations and pioneering simulation studies scaled
the partial charges by half to effectively include electronic
polarizability \cite{jonsson86,egberts94}. Also approaches that
explicitly include electronic polarizability has been introduced, but
this has turned out to be practically complicated \cite{chowdhary13}.
In this work we effectively include electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. The approach is more sophisticated version of
early studies with scaling to half was used \cite{jonsson86,egberts94},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13}.

%is a potential source of artefact, which could lead to overbinding
%of cations. 
%Almost all lipid bilayer simulations reported in the literature 
%are done with force fields that represent interactions between atoms
%with pairwise additive empirical potential and exclude the electronic 
%polarizability \cite{SOME_review, FF_papers}.

According to ECC, electronic
polarizability can be included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
The value measured for water, $\epsilon _{el} = 1.78$, gives 
a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}, which has been
successfully used to improve the performance of ion force fields \cite{kohagen14,kohagen16,??}.

The main goal of this work is to apply ECC correction to accurately 
describe the lipid headgroup response to Na$^+$ and Ca$^{2+}$ concentrations
in MD simulation models when compared with NMR data \cite{catte16}. 
The Lipid14~\cite{dickson14} force field parameters were used as a starting point,
because their response to bound ions was most realistic against NMR data
(see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and headgroup structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.

While the scaling factor of $f_q = 0.75$ for ions in water is well justified and shown
to improve model performance against scattering data \cite{kohagen14,kohagen16,??},
it is not clear if the same factor should be used for lipid headgroups \todo{, because ??}.
Here we estimated the correct scaling factor for Lipid14 model by using
so called implicitly polarized charges (IPolQ) \cite{ipolq2013},
which are the average of partial charges given by RESP~\cite{RESP_paper} calculations in
vacuum and in implicit solvent.
%In order to implicitly capture solvent induced polarization 
%in models with fixed charges, 
%we shall use the average of partial charges 
%from both environments (i.e. vaccum and solvent),
%so called implicitly polarized charges (IPolQ) \cite{ipolq2013}. 
The partial charges in original Lipid14 parametrization were derived by fitting the electrostatic 
potential to its model quantum chemistry representation (RESP\cite{RESP_paper}) in vacuum,
while charges from calculations in an implicit solvent vary most for the polar moieties \cite{maciejewski14}. 
Here we estimate the appropriate scaling factor for lipid hedgroups and glycerol backbone
by comparing IPolQ charges (average of RESP calculation in vacuum and in implicit solvent) to
the RESP results in vacuum. By using the results reported in \cite{maciejewski14} we conclude
that the IPolQ charges can be achived from vacuum calculations by scaling roughly with the
factor of $f_q=0.8$, which is the used to derive ECC-corrected Lipid14 model.
\todo{SAMULI: please check that this is correctly descrived.}
%By taking the charges of oxygen atoms in vacuum and implicit water solvent for POPC from \cite{maciejewski14}, 
%we represent the IPolQ charges in the electronic continuum correction of the Lipid14 model
%by increasing the scaling factor $f_q$ to $0.8$. 
\footnote{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  I use 0.80 as an approximate round value. The use of 0.78---.79 might be more appropriate, though.}

Here we apply ECC correction with the proposed scaling factor only to the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
highly optimized in Lipid14 and other lipids models and
give generally a good description for hydrophobic part of lipid bilayers
in various conditions \cite{ollila16},
in contrast to glycerol backbone and headgroup regions which require
some improvement in all available lipid models \cite{botan15}. 
%To minimize the detuning of the highly optimized hydrocarbon
%parameters, 
%Because the biologically relevant cations like \ce{Na^+} and \ce{Ca^{2+}}
%are found to interact only with the phospholipid head group,
%and the accuracy of alkyl sidechains in Lipid14 model are satisfactory, 
%we exclude alkyl tails from our modifications. 
The scaling of partial charges in the polar region was found to decrease
area per molecule to ??\todo{find the value}, which is significantly smaller than the
experimental value (\cite{}) and the original Lipid14 values (\cite{})\todo{Add the values}.  
The decrease of area was found to arise from lower hydration of the lipid headgroup region,
%\todo{SAMULI: Did you analyze the effect of this to hydration or did we only speculate?
%JOE: I found a systematic decrease of water density in the heagroup region even if I kept APL constant.}
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The solvation free energy can be decreased
and hydration increased by reducing the radius of beads by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms.
%This allows water molecules to approach closer to lipid atoms and
%have stonger electrostatic interactions with them.
%\todo{SAMULI: This effect may have an official name. In that case we should mention it.}
Here we reduce $\sigma$ with a scaling factor of $f_\sigma = 0.89$ for the same atoms for which
charges was scaled. This increased the area per molecule back
to a level close to experimental and original Lipid14 values. 
%\todo{Is there some justification for the below sentence?} 
%In line with the electronic continnum correction,
%the scaling factor for sigma parameters, $f_\sigma$,
%has to lay between $f_q$ and 1.

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept}
Ion binding was compared between experiments and simulations by 
using lipid headgroup order parameters and ''electrometer concept'' 
introduced by Seelig et al. \cite{seelig87,catte16}.
The concept is based on the experimental observation that the 
order parameters of $\alpha$ and $\beta$ carbons in lipid headgroup
(see Fig. \ref{??} 
\todo{Figure with chemical structure and labeling to be added}) 
are proportional to the amount of bound charge
in lipid bilayer \cite{seelig87}.
More recent analysis included also
the order parameter signs and concluded that the order parameters  
decrease with bound positive charge and increase with bound negative 
charge \cite{ollila16,catte16}. The observations are rationalized 
as a change of lipid headgroup dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
Order parameters for C-H bonds can be accurately measured for each
C-H bond of lipids by using H$^2$ NMR or $^{13}$C NMR techniques and
calculated from MD simulations directly from definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

The change of the headgroup order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Combination of atomic absorption spectra, $^2$H NMR experiments
and information about order parameter signs gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.

Recent work published by the NMRlipids project showed that the concept works
qualitatively also in simulations and can be used to compare ion binding
affinitity between simulations and experiments~\cite{catte16}. However,
it also turned out that also the sensitivity of the order parameters response
to bound charge has to be quantified more accurately for quantitative comparison
of binding affinity. In this work we first we first quantify the response of
headgroup order parameters against experimental data from cationinc surfactants
embedded in PC bilayer~\cite{scherer89}. Essentially all surfactants locate in
lipid bilayers in these experiments, thus the amount of bound charge is known
and can be plotted against order parameter change.

@@

\subsection{Comparison of lipid bilayer structure to experimental data}
Lipid bilayer structure without ions was validated against NMR experiments
by order parameters for C-H bonds and x-ray scattering experiments by using
form factors. The former validates the structures sampled by individual
lipid molecules in simulations with segmental resolution, while latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}. Form factor was calculated from simulations
as 
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering lenght,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
Water molecules were described by OPC3 model \cite{Izadi16}.
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987} and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion model by Dang \cite{smith94,chang1999,dang2006}. 
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

\begin{table}[btp]
  \caption{Simulation parameters}
  \label{tbl:mdpar}
  \begin{tabular}{ll}
    simulation property & parameter   \\
    \hline
    time-step           & 2~fs         \\
    equilibration time  & 100~ns  \\
    simulation time     & 200~ns  \\
    temperature         & 313~K       \\
    thermostat          & v-rescale  \cite{bussi07}   \\
    barostat            & Parrinello-Rahman, semi-isotropic \cite{parrinello81} \\
    long-range electrostatics & PME  \cite{darden93}  \\
    cut-off scheme      & Verlet \cite{Pall13}      \\
    Coulomb and VdW cut-off & 1.0~nm \\
    constraints         & LINCS, only Hydrogen atoms \cite{hess97} \\
    constraints for water & SETTLE  \cite{miyamoto92} \\
    \hline
  \end{tabular}
\end{table}

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
The code AmberTools \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
These systems were ran with original Lipid14 POPC model and with ECC-corrected model.
\todo{Describe the changes in surfactant model used with ECC-corrected lipids.}

@@

\subsection{Lipid bilayers without ions}
The scattering form factors, NMR order parameters and area per lipid calculated  
using the ECC-lipids model for POPC are compared to experiments and original
Lipid14 results in Fig. \ref{simVSexpNOions} and in Table~\ref{tab:apls}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.
The structural quality is comparable to the state of art lipid models available in literature \cite{ollila16},
thus we conclude that the ECC-lipids model reproduces the lipid bilayer structure
in liquid disordered phase with similar accuracy to other available models.

\begin{figure}[]
  \centering
  \includegraphics[height=8.5cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \includegraphics[width=9.0cm]{../Fig/ipython_nb/Headgr_OPs_exp-L14-ECCL17_q80_sig89.png}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Headgroup and glycerol backbone order parameters from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{Add acyl chain order parameters, POPC chemical structure} \\
  \todo{We should run a simulation for the system withtout ions with the
    same (or close) temperature as the experimental data. The NMR data is at 300K
    so that would be good. I think that this would be also close enough for the scattering data at 303K.
  }
\end{figure}
